Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-07-03 manuscript/manuscript.tex
Top to bottom review in attempt to summarize page count.
Edited and tables/charts removed.  Currently at 22 pages.

==========REVIEW BELOW HERE===========

%\subsection{Issues with Community Boards}
\label{sec:communityboards}
Community Boards are an important aspect of NYC government. They are the 
most local, grassroots form of City government, and a vital connection 
between communities and elected officials and City agencies. Community 
Boards are used as a measure of City services throughout the five 
boroughs. In the 2022-2023 dataset there are 27,276 invalid community\_board 
entries which represents 0.43\% of non-blank data. The distribution 
of invalid Community Boards by Agency is not consistent with 
the overall SR Agency distribution. This indicates that there are likely 
specific issues at some key Agencies, in this case Taxi \& Limo 
Commission (TLC), Parks \& Recreation, etc.

%\subsection{Issues with Community Districts}
\label{sec:communitydistrict}
Another of the ``computed'' fields is community\_districts. Community Districts 
are the boundaries for the Community Boards, but unlike how 
Community Boards are an arm of the government of New York City, 
the Community District is used by the Department of City Planning 
(DCP) for purposes of environmental, socio-economic, and demographic 
purposes. It is a geographical division rather than a local 
government division. Due to how the community\_district data 
is formatted, it is not possible to establish validity. However, 
it is possible to determine that the dataset contains 72 unique 
entries, while there are only 59 valid Community Districts.

%\subsection{created\_date and closed\_date(s) -- Negative Duration}
\label{sec:negativeduration}
For the created\_date and closed\_date fields, you might 
expect that these two fields are automatically populated by the 
SR application software. Thus the created\_date is associated with setting an SR status 
to ``new'' or ``open''. And similarly when an SR's 	status is 
changed to ``closed''  the closed\_date would be 
populated. Unfortunately, that does not appear to be the case as
there are several anomalies with the various date fields.

\begin{itemize}
	\item SRs with a closed\_date that occurs before the created\_date.
 	\item created\_date(s) and closed\_date(s) in the far distant past.
 	\item created\_date(s) and closed\_date(s) that match to the second.
	\item A large number of SRs closed and created exactly at midnight 
	or exactly at noon. 
\end{itemize}

These date fields are important because citizens, NYC Government Officials, 
and NYC Agencies use these dates to measure the 
responsiveness for providing services. The measure of ``duration'',	 
of the response time of a 311 call is a closely 
watched, carefully scrutinized metric, both in terms of overall performance 
and with a geographical area. Duration, while not a field in the dataset, 
it is easily computed by closed\_date - created\_date.

Closed-before-Created:  There are 12,251 SRs that are closed before they 
were created, thereby generating a nonsensical ``negative duration''. 
While this is a small percentage overall (0.2\%) it can have significant impact 
on response time analysis. Here is a sample:

@@

The largest errors are shown ``excluding extreme negative values''. We found 
eight SRs with extremely large negative durations (\textless{} -730), all containing 
an entry of ``1900-01-01'' as the closed\_date, which generates extremely large 
negative durations exceeding 44,601 days (122 yrs). That large of an anomaly 
can skew statistical results, even though the number of occurrences is 
small. According, these  SR rows are removed from the analysis. These 
extremely large negative duration SRs are all from the 
Department of Homeland Services (DHS) A visualization of the 
negative-duration SRs by Agency shows that this negative-duration 
issue is a DOT problem, where 95\% of these types of errors occur.

@@

This violin chart shows the broad spread of negative-duration SRs, albeit with 
the extremely large negative-duration SR removed. While there are few 
outliers, the magnitude of the negative-durations is troubling and can 
produce bizarre analytical results.

@@

%\subsubsection{Case Study: Homeless Person Assistance}
\label{sec:homlessassistance}
\textbf{Scenario:} The Dept. of Homeless Services (DHS) wants to 
know ``How quickly are 311 calls for ``Homeless Person Assistance
resolved over the past two years?'' This is a typical request made by 
both the public and City government. It's a performance metric along 
the lines of ``How quickly is my Agency responding to ``critical'' 
requests, and does that performance vary by Borough, Zip Code, etc. 
As an analyst, this appears to be a routine effort:

\begin{itemize}
    \item Select data from 2022-2023 and filter the data by 
    complaint\_type = ``Homeless Person Assistance'' (yields 55,000 SRs)
    \item Compute a new field ``duration'' (closed\_date – created\_date)
    \item Take an average of the ``duration'' field == \textbf{Answer:  -4.8 days}  
\end{itemize}

Clearly that answer is nonsensical. How did such a simple task result 
in an absurd answer? The answer lies in the computation of the ``duration'' 
field. As it turns out, there are eight DHS SRs that have a closed\_date 
of ``1900-01-01''. Each of those SRs creates a negative duration of -44,602 
days (-122 years). So just those eight SRs with extreme negative duration 
values is enough to drive the average of the 75,000 
Homeless Assistance SRs to a negative value, clearly incorrect.

In this case, the median is a better measure of central tendency than the average. 
And the median of the duration field is  0.2 days (approx 5 hrs).

@@

%\subsection{created\_date and closed\_date(s) --  Zero Duration}
\label{sec: zeroduration}		
A more serious and prevalent problem with closed\_date(s) 
occurs when the closed\_date and created\_date are exactly the same -- to 
the second -- which accordingly creates a \textbf{zero duration'}. Again 
this is nonsensical, and the presence of such zero durations can again 
severely distort analytics. There are 191,141 such SRs representing 3.1\% of 
all non-blank data. As shown in the chart 	below, 99\% of the 
zero duration SRs appear to predominately occur in five 
Agencies (Dept of Mental Health \&Hygiene, Dept of Transportation, 
Dept of Business, Dept of Sanitation, and Dept of Environmental Protection.) 
This is not in-line with the overall Agency distribution of SRs and indicates an 
Agency-specific issue, and the solution likely lies at those Agencies.

@@

%\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
\label{sec:midnightandnoon}
Another problem discovered with the created\_date and 
closed\_date fields - there appears to be an unusually large number of SRs 
created or closed at exactly midnight (00:00:00) and exactly noon (12:00:00), 
to the second. The distribution of SR creation and closure largely follows the 
work-day clock with many SRs created during day-light hours, and 
fewer SRs 	created at night and the early hours of the morning. However, 
here there appears to be significantly greater numbers of SR closed exactly 
at midnight and exactly at noon, as well as a significant number of SRs 
created exactly at noon. Observations during this 2-year period include:

\begin{itemize}
	 \item There were 99,779 SRs created exactly at noon (12:00:00)
	 \item There were 235,347 SRs closed exactly at midnight (00:00:00)
	\item There were 105,505 SRs closed exactly at noon. 
\end{itemize}

One other way to visualize this anomaly is to look at the busiest day 
during this 2-year period (Friday, 2023-09-29). Then aggregate 
the created\_date(s) by minute (with seconds equal to zero), providing
a minute-by-minute look at SR creation. Here we see a clear 
spike at exactly noon (12:00), well beyond the 3\textsigma line.

@@

As before, let's visualize this anomaly by looking at the closed\_date(s) 
on the busiest day during this 2-year study period (Friday, 2023-09-29). 
We aggregate closed\_date(s) by the minute (with seconds equal to zero). 
The chart highlights anomalies with spikes at 00:00:00 and 
12:00:00 respectively.

@@

These unusual patterns of created \& closed SRs at exactly the hours 
of midnight and noon likely indicates the presence of a bulk create/close 
software process that perhaps automatically ``closes'' 
(or ``creates'') a large number of SRs with a provided time-stamp of 
midnight (00:00:00) or noon (12:00:00). Again, such behavior will
distort the duration of these SRs. Looking at the distribution 
by Agency for the ``closed-exactly-at-midnight'' 
shows that \textgreater90\% of these suspect SRs come from just two 
Agencies - Dept. of Buildings (DOB) and Dept. of Sanitation (DSNY).

@@

However, the ``SR closed at noon'' visualization shows that  a single 
Agency, DSNY, is responsible for \textgreater99\% of the 
SRs ``closed-exactly-at-noon''.

@@

%\subsection{resolution\_action\_update\_date}
\label{sec: resolutionaction}
When an SR is updated, by any means on any field, the 311 software 
automatically populates the resolution\_action\_update\_date. 
Analyzing that field it becomes apparent that some of the SR updates 
are happening long after the SR is closed. There are a 
total of 7460 SRs that are updated \textgreater30 days and \textless{}730 
days after the closed\_date (to avoid infeasible dates such as 1900-01-01).
It is not known if this is normal behavior or an area that may 
require further investigation.

@@

\section{Accuracy and precision}
\label{sec:precision}
An area where the question of precision vs. accuracy arises is
with the Latitude and Longitude fields. Both the Longitude and 
Latitude fields are expresses as a 14-decimal 
number, e.g. 40.86769186022511. Also the Location field which is a 
straight concatenation of latitude and longitude). Given that 1 
degree of latitude at the equator is equal to 111.044736 kilometers, 
the ``1'' at the end of that number represents approximate 1.1104 nanometer 
or 1/1,000,000,000 of a meter. For reference a DNA molecule is 
approximately 2nm in width. Clearly the representation of 
the Latitude and Longitude fields are a classic case of 14-digit 
precision, but limited accuracy.

\section{Redundant \& Duplicate fields}\label{sec:duplicates}
During this analysis, several redundant fields were observed. These 
fields should be examined further for possible consolidation.

%\subsection{latitude \& longitude and the location fields}
\label{sec:latlong}
The location field is a pure concatenation of the latitude 
and longitude fields, with a comma and parenthesis 
added. The inclusion of the location field seems questionable 
as the data is arguably more difficult to extract than the 
two specific fields, especially for software programs. Example:

\begin{itemize}
	\item  latitude: 40.768456429488
	\item  longitude: -73.9575661888774
	\item  location: (40.768456429488, -73.95756618887745)
\end{itemize}

%\subsection{borough and park\_borough fields}
\label{sec:parkborough}
These two fields are 100\% matches; fully redundant.

%\subsection{borough and borough\_boundaries fields}
\label{sec:boroughboundaries}
The borough field is a text field for the five New York City boroughs (Manhattan, Queens, 
Brooklyn, etc.). The borough\_boundary field is numeric with values 
from 1-5 corresponding to the five boroughs (Staten Island, Brooklyn, 
Queens, Manhattan, Bronx). When the borough\_boundary field is 
translated and compared to the borough field, there is a 
98.3\% match. And that is the problem with two ``near-duplicate'' 
fields; which field is correct when they disagree?

%\subsection{borough and taxi\_company\_borough fields}
\label{sec:taxicompanyborough}
The two fields borough and taxi\_company\_borough might appear to be 
related. In fact, despite the names, the two data fields are nearly 
completely different with only 0.05\% matching. The mismatch is so 
high that it suggests the two fields are used very differently. This field 
is used exclusively by the Taxi \& Limo Commission which governs 
taxis and other cars for hire.

 %\subsection{incident\_zip and zip\_codes fields}
 \label{sec:zipcodes}
 The zip\_code field is one of the six computed fields and its validity 
 has been explored in Section \ref{sec:zip-codes} and found to be 
 lacking with 55.55\% of the entries invalid according to the 
 USPS, while the companion zipcode field incident\_zip field had 
 an accuracy rate of 99.93\%.

 %\subsection{police\_precinct and police\_precincts fields} 
 \label{sec:police} 
Both the police\_precinct and police\_precincts are among the computed 
fields not cited in the Data Dictionary. Their usage however, is easy to 
discern. These two fields are near-duplicates with 99.94\% of the entries 
matching. Since the source of the fields and the underlying computational 
process are not know, it remains a challenge to determine which 
field is the ``right'' field to use. (The validity of these fields is 
covered in Section \ref{sec:police-precincts}.)

% \subsection{agency and agency\_name fields}
 \label{sec:agencyname}
 While not duplicates, these fields have a 1:1 correspondence.  The Agency 
 field contains abbreviations for the City agencies such as NYPD, DOT, 
 HPD, etc. The agency\_name field contains the full name of 
 the various organizations: New York Police Department, Department 
 of Transportation, Department of Housing Preservation and 
 Development, and so on. It seems redundant to include both.

%\subsection{landmark and street\_name}
\label{sec:landmark}
The landmark field is listed in the Data Dictionary as ``Can refer to 
any noteworthy location, including but not limited to, parks, 
hospitals, airports, sports facilities, performance spaces, etc.'' Our analysis
found that was not the case. To be sure many of the entries in 
the landmark field do contain landmark names, e.g.``Pennsylvania 
Station'', ``La Guardia Airport'', etc., the vast majority 
of the entries are street names, e.g. ``Fenton Avenue'', ``Steinway 
Street'', ``Broadway''. These entries are similar to those in the 
street\_name field. And in fact, street\_name and landmark 
match 62\% of the time. While a high enough percentage of matching 
to be indicative of duplicate usage. Even the non-matches 
(excluding blanks) would appear to be matches except for minor 
spelling and nomenclature changes, e.g. ``NINTH AVE'' \& ``9 AVE''.

@@

%\subsection{cross\_street\_1 \& intersection\_street\_1 and cross\_street\_2 
\& intersection\_street\_2}
\label{sec:street1}
There are two sets of street pairs in the dataset in addition to 
the complaint address:

\begin{itemize}
	\item cross\_street\_1
	\item cross\_street\_2
	\item intersection\_street\_1
	\item intersection\_street\_2
\end{itemize}

These two pairs of streets are used to help identify the location of the 
reported incident, as it is common in New York City to provide the 
street address and the cross street, as in ``24 W 90th street between 
Columbus and Amsterdam''. For the majority of data rows, both 
pair of fields are populated and are duplicates; 88\% of 
cross\_street\_1 matches intersection\_street\_1. Similarly with
cross\_street\_2 exactly matches intersection\_street\_2.

\textit{(Note: Address standardization was applied to the street 
addresses using the R \emph{campfin} package. This was done 
to prevent generating a non-match between 
streets such as ``240 E 69\textsuperscript{th} ST'' and ``240 E 
69\textsuperscript{th} Street'' which are clearly the same address
but with formatting and spelling differences.)}

In addition to the matching/non-matching values, we encountered a number 
of \emph{near-matches}, cases where the 
cross\_street and intersection\_street are nearly identical, 
e.g. HARMON DR and HARMON RD. To identify these 
near-matches we measured the \href{https://en.wikipedia.org/wiki/Hamming_distance}
{Hamming Distance} between the two fields. The Hamming Distance is a 
measure of how many letters have to be changed in order for the two 
fields to match. For this analysis, we chose a Hamming Distance of 2. Here 
are the results for both cross\_street and intersection\_street.

@@

 %\subsection{Shrinking file size by removing duplicates}
 \label{sec:filesize}
By removing duplicate and ``near-duplicate'' fields, it is possible to 
shrink the file size by 12.3\% which for this dataset equates to a 
reduction of 395 Mb. A smaller file size means faster downloads, 
less storage impact, as well as simplifying data analysis efforts.

Below is a list of duplicate and near-duplicate fields. There are 
some challenges with these proposed deletions, mostly with 
the ``near duplicate'' fields.  So while the park\_borough is 
a complete duplicate of the borough field, the 
borough\_boundary field is only 98.3\% duplicate. Admittedly, 
loosing 1.7\% of the data in those two fields is quite small, however
for this large dataset that amounts to 110, 715 
non-matching occurrences. The authors  propose 
deleting the following fields from the data set.

@@

Additionally, it would be worth investigating if certain data fields are 
in-fact useful, as the population of these fields is very, very scarce. For 
example, the taxi\_company\_borough field is 99.94\% 
blank. A similar situation exists for the road\_ramp field 
(99.75\% blank), the vehicle\_type field (99.71\%blank), and 
due\_date which is 99.62\% blank. There are several fields that are 
mostly blank: bridge\_highway\_direction, bridge\_highway\_name, 
bridge\_highway\_segment, and taxi\_pick\_up\_location...all 
of which have \textgreater99\% blanks. Most of those fields 
appear to used exclusively by the Taxi \& Limo Commission (TLC).

\section{Data Dictionary Observations} 
\label{sec:datadictionary}
During this analysis, it became apparent that the  Data 
Dictionary could use an update. We found several discrepancies 
between the Data Dictionary and the actual data which can lead 
analytic efforts astray.

\begin{itemize}
	\item While the Open Data portal lists the data types of the 
	various fields, the Data Dictionary does not. Nor does it provide 
	much information on the specifics of various fields, such 
	as constraints or domain of legal values. For example, the 
	incident\_zip field is specified as ``text''. That's probably not the 
	best definition, perhaps	``categorical'' indicating that the data can 
	contain only numeric characters and is not subjected to arithmetic 
	operations.

	\item Many of the domain of legal values are incomplete or 
	inaccurate. For example, the Data Dictionary indicates that 
	the ``status'' field has values of \textit{assigned, canceled, closed}, 
	or \textit{pending}. However, we found additional status 
	values of: \textit{in progress, started}, and \textit{unspecified}. 
	However, we found no SRs with a status of \textit{canceled}. 
	Similarly, the address\_type field is indicated to have 
	values \textit{address, blockface, intersection, latlong}, and 
	\textit{placename}. We discovered the additional values of 
	\textit{bbl} and \textit{unrecognized}. Additonally, no values of 
	\textit{latlong} were observed. Other fields (facility\_type, 
	vehicle\_type, taxi\_pick\_up\_location, road\_ramp, city) 
	suffer similar inaccuracies in their doman of values.

	\item As noted previously, there are six \textit{@computed} fields that
	are present in the data, but not described in the Data Dictionary.
\end{itemize}

\section{Recommendations} 
\label{sec:recommendations}
Here are recommendations as to how to resolve these issues.

\begin{itemize}
	\item First, and most importantly, the Open Data Team will need 
	to undertake a dedicated effort to identify these 
	data anomalies, confirm the veracity of the issue, 
	and engage with the appropriate Agency Open Data 
	Representatives to resolve these issues. That effort will not 
	be easy, fast, or cheap and in some cases software changes will likely 
	be required. Coordination and mutual approval of any 
	changes will be key, and solid support from senior leadership 
	will be required.

	\item Secondly, standards and protocols will need to be 
	identified, agreed to, and commitment obtained. For example, 
	what are the domain of legal values for certain fields?

	\item Many of the data integrity issues could potentially be 
	resolved through rule enforcement by software. For example, the 
	software could easily prevent invalid values, such as zip 
	codes. Similarly for other data issues. One key area are is the 
	created\_date and closed\_date which are frequently cited 
	when determining responsiveness of various Agency services.

	\item The large spikes in SR closure and creation occurring at 
	midnight and noon, almost certainly point to an automated 
	process from an external Agency integration. Such processes again distort 
	the accurate capture of an SR duration.

	\item Investigate moving additional Agencies to the core 
	311 software system where enforcement of data rules would 
	presumably be more consistent.

>>>>

%\subsection{Issues with Community Boards}
\label{sec:communityboards}
Community Boards are the most local, grassroots form of 
City government; a vital connection between communities and 
elected officials and City agencies. Community Boards are often
used as a measure of equity of City services throughout the five 
boroughs. In the 2022-2023 dataset there are 27,276 invalid community\_board 
entries which represents 0.43\% of non-blank data. The distribution 
of invalid Community Boards by Agency is not consistent with 
the overall SR Agency distribution. This distribution indicates that 
there are likely issues at key Agencies, in this case Taxi \& Limo 
Commission (TLC), Parks \& Recreation, etc.

%\subsection{Issues with Community Districts}
\label{sec:communitydistrict}
Another of the ``computed'' fields is community\_districts. Community Districts 
are the boundaries for the Community Boards, but differ as they
are used by the Department of City Planning (DCP) for purposes 
of environmental, socio-economic, and demographic 
purposes. It is a geographical division rather than a local 
government division. Due to how the community\_district data 
is formatted, it is not possible to establish validity. However, 
it is possible to determine that the dataset contains 72 unique 
entries, while there are only 59 valid Community Districts.

%\subsection{created\_date and closed\_date(s) -- Negative Duration}
\label{sec:negativeduration}
For the created\_date and closed\_date fields, you might 
expect that these two fields are automatically populated by the 
SR application software; associated with setting an SR status 
to ``new'' or ``closed''. Unfortunately, that does not appear 
to be the case as there are several anomalies with the various date 
fields, including:

\begin{itemize}
	\item SRs with a closed\_date that occurs before the created\_date.

 	\item created\_date(s) and closed\_date(s) in the far distant past.

 	\item created\_date(s) and closed\_date(s) that match to the second.

	\item A large number of SRs closed and created exactly at midnight 
	or exactly at noon, to the second
\end{itemize}

Citizens, NYC Government Officials, and NYC Agencies use 
these dates to measure the responsiveness for providing 
services. In particular, the of ``duration'' of the SR 
is a carefully scrutinized metric used to measure the responsiveness
of an Agency or service. Duration, while not directly present 
in the dataset, is easily computed by (closed\_date - created\_date).

Closed-before-Created:  There are 12,251 SRs that are closed before they 
were created, thereby generating a nonsensical ``negative duration''. 
While this is a small percentage overall (0.2\%) these errors can 
have significant impact on response time analysis. 
Here is a sample:

@@

As noted, the largest errors are shown ``excluding extreme negative values''. We found 
eight SRs with extremely large negative durations (\textless{} -730), all originating
with DHS and all containing an entry of ``1900-01-01'' as the closed\_date. This 
generates extremely large negative durations exceeding 44,601 
days (122 yrs). That large of an anomaly can significantly skew statistical 
results, even though the number of occurrences is small. According, these 
SR rows are removed from the analysis. Excluding these extreme negative
values, a visualization Agency indicates that the negative-duration 
issue is a DOT problem, where 95\% of these types of errors occur.

@@

This violin chart shows the broad spread of negative-duration SRs, albeit with 
the extremely large negative-duration SR removed. While there are few 
outliers, the magnitude of the negative-durations is troubling and can 
produce bizarre analytical results as will be shown below.

@@

%\subsubsection{Case Study: Homeless Person Assistance}
\label{sec:homlessassistance}
\textbf{Scenario:} DHS wants to know ``How quickly have 311 SRs
for ``Homeless Person Assistance'' been resolved over the 
past two years?'' This is a typical request made by 
both the public and City government; analgous to ``How quickly 
is my Agency responding to ``critical'' requests, and does 
that performance vary by Borough, Zip Code, etc. Here is a
sample analysis:

\begin{itemize}
    \item Select data from 2022-2023 and filter by complaint\_type = 
    ``Homeless Person Assistance'' (this yields 55,000 SRs)

    \item Compute the ``duration'' (closed\_date – created\_date)

    \item Take an average of the ``duration'' field == \textbf{Answer:  -4.8 days}  
\end{itemize}

Clearly that answer is nonsensical. How did such a simple task result 
in an absurd answer? The answer lies in the computation of the ``duration'' 
field. It turns out, there are eight DHS SRs that have a closed\_date 
of ``1900-01-01''. Each of those SRs creates a negative duration of -44,602 
days (-122 years). Just those eight SRs is enough to drive the 
average of the 75,000 Homeless Assistance SRs to a negative 
value; clearly incorrect. In this case, the median is a better measure 
of central tendency; the median is  0.2 days (approx 5 hrs).

@@

%\subsection{created\_date and closed\_date(s) --  Zero Duration}
\label{sec: zeroduration}		
A more prevalent problem with closed\_date(s) occurs when 
the closed\_date and created\_date are exactly the same -- to 
the second. Accordingly, this creates a \textbf{zero duration'}, again 
nonsensical. There are 191,141 such SRs representing 3.1\% of 
all non-blank data. As shown below, 99\% of these zero duration 
SRs occur in five Agencies:  (Dept of Mental Health \&Hygiene (DHMH), Dept 
of Transportation (DOT), Dept of Business (DOB), Dept of Sanitation 
(DSNY), and Dept of Environmental Protection (DEP). This distribution
is does not mirror the overall Agency distribution of SRs indicating an 
Agency-specific issue.

@@

%\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
\label{sec:midnightandnoon}
Another problem discovered with the created\_date and 
closed\_date fields: there are an unusually large number of SRs 
created or closed at exactly midnight (00:00:00) and exactly noon (12:00:00), 
to the second. Normally, distribution of SR creation and closure 
largely follows the work-day clock with many SRs created during 
day-light hours, and fewer SRs 	created at night and the early hours 
of morning. However, there are significantly greater numbers 
of SR closed exactly at midnight and noon, as well as a significant 
number of SRs created exactly at noon. Observations include:

\begin{itemize}
	 \item There were 99,779 SRs created exactly at noon (12:00:00)

	 \item There were 235,347 SRs closed exactly at midnight (00:00:00)

	\item There were 105,505 SRs closed exactly at noon. 
\end{itemize}

A unique way to visualize this anomaly is to examine the busiest day 
during this 2-year period (Friday, 2023-09-29). Then aggregate 
the created\_date(s) by minute (with seconds equal to zero), providing
a minute-by-minute look at SR creation. Here we see a clear 
spike at exactly noon (12:00), well beyond the 3\textsigma line.

@@

As before, let's visualize this anomaly by looking at the closed\_date(s) 
on the busiest day (Friday, 2023-09-29). Aggregate closed\_date(s) by 
the minute (with seconds equal to zero). This highlights spikes 
at 00:00:00 and 12:00:00 respectively.

@@

These unusual patterns of created \& closed SRs at exactly the hours 
of midnight and noon likely indicates the presence of a bulk create/close 
software process that perhaps automatically ``closes'' 
(or ``creates'') a large number of SRs with a generated time-stamp of 
midnight (00:00:00) or noon (12:00:00). Again, such behavior distorts
 the duration of these SRs. The distribution by Agency for 
 the ``closed-exactly-at-midnight'' SRs indicates that \textgreater90\% of 
 these suspect SRs come from just two Agencies: DOB and DSNY.

@@

A further visualization reveals that  a single Agency, DSNY, is 
responsible for \textgreater99\% of the SRs ``closed-exactly-at-noon''.

@@

%\subsection{resolution\_action\_update\_date}
\label{sec: resolutionaction}
When an SR is updated, the 311 software automatically 
populates the resolution\_action\_update\_date. Some of 
those SR updates are happening long after the SR is 
closed. There are a total of 7460 SRs that are updated \textgreater30 days 
after the closed\_date (and \textless{}730 days to avoid infeasible 
dates such as 1900-01-01). It is not known if this is normal behavior
 or an area that requires further investigation.

@@

\section{Accuracy and precision}
\label{sec:precision}
An question of precision vs. accuracy arises with the Latitude 
and Longitude fields. Both Latitude are 
expressed as a 14-decimal number, e.g. 40.86769186022511. Given 
that 1 degree of latitude at the equator is equal to 111.044736 
kilometers, the ``1'' at the end of that number represents 
approximately 1.1104 nanometers (1/1,000,000,000 of a meter). For 
reference a DNA molecule is approximately 2nm in width. Clearly 
the representation of the Latitude and Longitude fields is a 
classic case of 14-digit precision, but limited accuracy.

\section{Redundant \& Duplicate fields}\label{sec:duplicates}
During this analysis, several redundant fields were observed which
should be examined further for possible consolidation.

%\subsection{latitude \& longitude and the location fields}
\label{sec:latlong}
The location field is a pure concatenation of the latitude 
and longitude fields with a comma and parenthesis 
added; 100\% duplicative. Thus the  location field seems questionable 
as the data is arguably more difficult to extract and use than the 
two individual fields. Example:

\begin{itemize}
	\item  latitude: 40.768456429488

	\item  longitude: -73.9575661888774

	\item  location: (40.768456429488, -73.95756618887745)
\end{itemize}

%\subsection{borough and park\_borough fields}
\label{sec:parkborough}
Borough and park\_borough: These two fields are 100\% matches; fully redundant.

%\subsection{borough and borough\_boundaries fields}
\label{sec:boroughboundaries}
Borough and borough\_boundary fields are a 98.3\% match. This 
is an example of the the problem with two ``near-duplicate'' 
fields; which field is correct when they disagree?

%\subsection{borough and taxi\_company\_borough fields}
\label{sec:taxicompanyborough}
Borough and taxi\_company\_borough fields that despite their 
names, are nearly completely different; only a 0.05\% match, suggesting
that these two fields are used very differently. The taxi\_company\_borough
 field is used exclusively by the TLC which governs all cars for hire.

 %\subsection{incident\_zip and zip\_codes fields}
 \label{sec:zipcodes}
 The zip\_code field is one of the six computed fields and its validity 
 was explored in Section \ref{sec:zip-codes} and found to be 
 lacking; 56\% of the entries are invalid. The companion 
 field, incident\_zip, had an accuracy rate of 99.93\%. Our analysis
 indicates that the computed zip\_code field should be eliminated
 until such time as accuracy issues can be addressed.

 %\subsection{police\_precinct and police\_precincts fields} 
 \label{sec:police} 
Both the police\_precinct and police\_precincts are among the computed 
fields not cited in the Data Dictionary. However, their usage is easy to 
discern; they represent the NYPD precinct that corresponds to the
incident address. These two fields are near-duplicates with 99.94\% of the entries 
matching. (The validity of these fields is 
covered in Section \ref{sec:police-precincts}.) Since the underlying 
computational process is unknown, we are unable to determine which 
field is ``more correct'' field. However having two fields with 99.9\% 
duplicity indicates that only one police precinct field is necessary.

% \subsection{agency and agency\_name fields}
 \label{sec:agencyname}
 Agency and agency\_name fields: While not duplicates per se, these fields 
 have a 1:1 correspondence.  The Agency field contains abbreviations 
 for the City agencies such as NYPD, DOT, HPD, etc. The agency\_name 
 field contains the full name of these organizations. It seems 
 redundant to include both as the abbreviations are commonly used
 and well understood.

%\subsection{landmark and street\_name}
\label{sec:landmark}
Landmark and street\_name: The landmark field is listed in the Data Dictionary as ``Can refer to 
any noteworthy location, including but not limited to, parks, 
hospitals, airports, sports facilities, performance spaces, etc.'' Our analysis
found that was not the case. To be sure many of the entries in 
the landmark field do contain landmark names, e.g.``Pennsylvania 
Station'', ``LaGuardia Airport'', etc., the vast majority 
of the entries are street names, e.g. ``Fenton Avenue'', ``Steinway 
Street'', ``Broadway''. These entries are very similar to the 
street\_name field with a 62\% match, enough to be indicative 
of duplicate usage. Even the non-matches (excluding blanks) 
appear to be matches except for minor spelling and 
nomenclature changes, e.g. ``NINTH AVE'' \& ``9 AVE''.

@@

%\subsection{cross\_street\_1 \& intersection\_street\_1 and cross\_street\_2 
\& intersection\_street\_2}
\label{sec:street1}
Cross\_street\_1 \& intersection\_street\_1 and cross\_street\_2 
\& intersection\_street\_2: There are two sets of street pairs in addition to 
the incident\address field. These pairs are:

\begin{itemize}
	\item cross\_street\_1 \& intersection\_street\_1

	\item cross\_street\_2 \& intersection\_street\_2
\end{itemize}

These two street pairs used to assist in identifying the address, as 
is common in New York City. These two street pairs
have 88\% duplicates where cross\_street\_1 matches 
intersection\_street\_1 and cross\_street\_2 
matches intersection\_street\_2. We believe only one of these
street pairs are necessary. However it continues to illustrate the 
problem of two similar data fields that largely, but not exactly, match. In
this case there are 12\% that do not match; which field to trust.

\textit{(Note: Address standardization was applied to the street 
addresses using the R ``campfin'' package. This prevents 
generating a non-match between locations such as 
``240 E 69\textsuperscript{th} ST'' and ``240 E 69\textsuperscript{th} Street'' 
which are clearly the same address save for formatting and spelling differences.)}

In addition to the matching values, we encountered a number 
of \textit{near-matches}, situations where the 
cross\_street and intersection\_street are nearly identical, 
e.g. HARMON DR and HARMON RD. To identify these 
near-matches we measured the \href{https://en.wikipedia.org/wiki/Hamming_distance}
{Hamming Distance} between the two fields. Hamming Distance is a 
measure of how many letters must be changed for the two 
fields to match. For this analysis, we chose a Hamming Distance of 2. Here 
are some sample results for both cross\_street and intersection\_street.

@@

 %\subsection{Shrinking file size by removing duplicates}
 \label{sec:filesize}
Shrinking file size by removing duplicates: By removing duplicate 
and ``near-duplicate'' fields, it is possible to 
shrink the file size by 12.3\% which for this large dataset equates to a 
reduction of 395 Mb. A smaller file size means faster downloads, 
less storage impact, as well as simplifying data analysis efforts.

Below is a list of duplicate and near-duplicate fields. There are 
some challenges with these proposed deletions, mostly with 
the ``near duplicate'' fields.  For example, while the park\_borough is 
a complete duplicate of the borough field, the 
borough\_boundary field is only 98.3\% duplicate. Admittedly, 
loosing 1.7\% of the data in those two fields is quite small, however
for this large dataset that small percentage still amounts 
to 110, 715 non-matching occurrences. The authors  propose 
deleting the following fields from the data set.

@@

Additionally, it would be worth investigating if certain data fields are 
in-fact useful, as the population of these fields is very scarce. For 
example, the taxi\_company\_borough field is 99.94\% 
blank. A similar situation exists for the road\_ramp field 
(99.75\% blank), the vehicle\_type field (99.71\%blank), and 
due\_date which is 99.62\% blank. Several additional several fields
are mostly blank: bridge\_highway\_direction, bridge\_highway\_name, 
bridge\_highway\_segment, and taxi\_pick\_up\_location; all 
have \textgreater99\% blanks. Most of these fields 
appear to used exclusively by the TLC.

\section{Data Dictionary Observations} 
\label{sec:datadictionary}
The Data Dictionary could use an update. We found several 
discrepancies between the Data Dictionary and the 
actual data which can potentially lead analytic efforts astray.

\begin{itemize}
	\item While the Open Data portal lists the data types of the 
	various fields, the Data Dictionary does not. Nor does it provide 
	information on the specifics of various fields, such 
	as constraints or the domain of legal values. For example, the 
	incident\_zip field is specified as ``text''. That's probably not the 
	best definition; perhaps	``categorical'' with notes indicating that 
	the data can contain only numeric characters and is 
	not subjected to arithmetic operations.

	\item Many of the domain of legal values are incomplete or 
	inaccurate. For example, the Data Dictionary indicates that 
	the ``status'' field has values of \textit{assigned, canceled, closed}, 
	or \textit{pending}. However, we found additional status 
	values of: \textit{in progress, started}, and \textit{unspecified}. 
	Additionally, we found no SRs with a status of \textit{canceled}. 
	Similarly, the address\_type field is indicated to have 
	values \textit{address, blockface, intersection, latlong}, and 
	\textit{placename}. We discovered the additional values of 
	\textit{bbl} and \textit{unrecognized}. Additionally, no values of 
	\textit{latlong} were observed. Other fields (facility\_type, 
	vehicle\_type, taxi\_pick\_up\_location, road\_ramp, city) 
	suffer similar inaccuracies in their specified domain of values.

	\item As noted previously, there are six \textit{@computed} fields that
	are present in the data, but not included in the Data Dictionary.
\end{itemize}

\section{Recommendations} 
\label{sec:recommendations}

\begin{itemize}
	\item First, and most importantly, the Open Data Team will need 
	to undertake a dedicated effort to identify these 
	data anomalies, confirm the veracity of the issue, 
	and engage with the appropriate Agency Open Data 
	Representatives to resolve these issues. That effort will not 
	be easy or fast, and in some cases software changes will 
	be required. Coordination and mutual approval of any 
	changes will be key, requiring solid support 
	from senior leadership.

	\item Secondly, standards and protocols will need to be 
	identified, agreed to, and commitment obtained. For example, 
	the domain of legal values, data accuracy, address normalization, etc..

	\item Many of the data integrity issues could potentially be 
	resolved through enforcement by software. For example, the 
	software could easily prevent invalid values, such as zip 
	codes. Similarly for other data issues. One key area are is the 
	created\_date and closed\_date which are frequently cited 
	when determining responsiveness of various Agency services; software
	could conceivably prevent illogical situations such as an SR closed 
	before it is opened.

	\item The large spikes in SR closure and creation occurring at 
	midnight and noon almost certainly point to an automated 
	process from an external Agency integration. Such processes distort 
	the accurate capture of an SR duration and should be eliminated.

	\item Investigate moving additional Agencies to the core 
	311 software system where data rules would 
	be more consistently enforced.
